A game-theory modeling approach to utility and attachment dynamics in biomedical research social networks

### 2016-11-15 manuscript_3.tex
Comentarios de Enrique added

\title{A game-theory modeling approach to utility
  and strength of interactions dynamics in biomedical research social networks}
%\title{Two games modeling approach to collaboration dynamics in biomedical
%research social networks}

@@

  What would happen if researchers in a community were given the same
  amount of resources and then set free to interact among each other?
  How would these resources be distributed in the population after
  many interactions? The final result may be determined by the
  researchers' social and political power, an by their
  prestige. Nevertheless, all these attributes depend, to a certain
  degree, on the research collaboration social network in which
  scientists are embedded. Here we develop a model based on the
  Prisoner's Dilemma implemented on a collaboration network. With this
  model we explore the distribution of resources and the dynamics of
  the strengthening or weakening of collaboration connections among
  partners, as scientists cooperate or defect in order to gain access
  to others' resources. The network is based on data about projects
  that were awarded with a grant. In this sense, we assume that
  resources are funds, equipment and time. We define Fitness as how
  much access a researcher has to the resources of his collegues.  We
  tested our simulation on the real biomedical research network and
  compared the results with an Erd\"{o}s-Reny\'i, a Watts-Strogatz
  small-world and Barab\'asi-Albert topologies. Different topologies
  display different fitness and connections strength
  distributions. Moreover, the distribution of fitness and connections
  strength in the researchers network is similar to that of
  Barab\'asi-Albert and Watts-Strogatz topologies, respectively. We
  believe that fitness distribution in the researchers network
  suggests that there are socio-cultural mechanisms governing the
  network that produce an asymmetric distribution of resources. The
  high distribution of strong connections might reflect some sort of
  subordination among researchers by which they are morally obliged to
  cooperate by the same socio-cultural mechanisms. The range around
  the threshold that regulates the decision to cooperate or defect
  according to the agent's historical balance between utility and
  strength of collaborative relationships and carrying capacity of the
  system is small, suggesting that there is a region in which a phase
  transition takes place from a population of cooperators to a
  population of defectors. Simulations like this may help to develop
  science policies to promote fair distribution of resources.

@@

\textbf{Notes:}

\begin{itemize}

@@

\item We change utility for fitness. Utility seems a better concept in order to
  reflect players' preferences.

@@

\item Choosing a strategy depends on the balance $\eta'$ equation --look at page
  9. This
  equation states that the probability of an agent to cooperate increases as: a)
  its utility increases --as it moves away from 0, or; b) The strength of the
  connection with its peers increases --as it moves away from 0, or; c) both
  increases.

\item{En la pagina 9 decimos que ``In order for the agent to choose to cooperate
  or defect, a degree of willingness to cooperate is assigned to each agent.'' No estoy
  seguro que ``degree of willingness to cooperate'' sea la mejor manera de
  decirlo, mi duda está en la palabra ``degree'', por lo que cambi\'e la
  oraci\'on a: The probability for an agent to cooperate or defect depends on a
number that referes to a historical balance between average utility and the average strength of
the connections with its neigbors. }

\end{itemize}

@@

After each game, the agent adds-up utility ($u$), which is the sum of the
pay-offs following the PD matrix. A pair of neighbors will add-up to the strength of
their connection ($w_{ji}$) as they coordinate or anti-coordinate, being $w$
also cumulative. We measure global utility and connection strength for the whole
network. Global utility $U$ is the sum of all individual utilities and global
strength of connections or $W$ is the sum of every pair of agents' links
$w$. Metaphorically, the strength of connection can 
be thought as some sort of ``trust''.\\

{\color{red} It should be noted that the same actions or behaviors work for
  both $u$ and $w$. There are two reasons for this decision in the design of the
model. The most general one is that we believe that in the real world,
actions such as cooperating and defecting affect the strength of the connection
among people. The second one is that we think that selfishly maximizing resources and
strenghting relationships are \emph{opposing forces} acting on the same set of
behaviors. The actions of an agent imply a trade-off in which defecting may
increase its utility at the expense of its collaborative relationships. If
collaborators have nourished their relationships, they might be strong enough to
endure occasional defection. Cooperating may build up relationships but it can be
expensive for the player.}

@@

The probability for an agent to cooperate or defect depends on a
number ($\eta$) that referes to a historical balance between average utility and
the average strength of the connections with its neigbors.   {\color{red} This is so
  because we assume that whatever the result in utility or strength of
  connection, as long as one of them increases, the player will be confident in
  the strategy followed so far}.\\

@@

We ran two different experiments. In the first we simulated different values of
carrying capacity $\nu$. With this experiment we were able to see how the number of
cooperators, utility, strength of connections among agents and the ratio of
shifting state population would change in the range of the carrying
capacity. For this experiment the states of the agents were the same at
initialization, for all values of the carrying capacity. Since the model is
deterministic, it will return the same result if run under the same
conditions.\\

@@

In this section we present the main results of the study, namely the topological
structure of the underlying network models, the dynamics of the games under
different parameters and network topologies and the distribution of utility and
of the strength of interaction resulting of playing the games in all the
different scenarios considered, including the real FOSISS network.\\

@@

Other results to report are those of the dynamics of different
variables as the carrying capacity $\nu$ changes. The  most salient
result is that for $\nu$ between $0.19$ and $0.24$, there is an
apparent phase transition in all different topologies and for all the
different variables. Nevertheless, it is worth noticing that the shape
of the phase transition is different according to the topology of the
network at stake. When $\nu$ is between $0.0$ and $0.2$, that is, when there is
no space or a very short one for suspiciousness, all agents cooperate, when
\textit{carrying capacity} is above $0.25$, all agents defect. \textit{Utility},
\textit{strength of interactions}, and \textit{changing state population}
replicate that same behavior for the same limits. \\

@@

Utility and strength of interactions dynamics under different $\nu$ emulate the
phase transition found before. In \textbf{Figures 3 and 4}, it can be observed
that there is a drop in utility and strength of interactions according to the
drop in the number of cooperators. Erd\"{o}s-Reny\'i and Barab\'asi-Albert
networks are quite similar in the way these variables fall in two steps, the
first one at $\nu \approx 0.2$ and the next one at $\nu \approx 0.23$. The fall
is sharper still in the Barab\'asi-Albert topology. Utility and strength of
interactions phase transition in Watts-Strogatz network is significantly more
staggered compared to the former networks. In the case of utility, there is a
region in the limits of $\nu \approx 0.25$ and $\nu \approx 0.3$, before utility
goes to $0$, in which it remains low but stable over time. In general, strength
of interactions follows the same pattern as utility but in the same $\nu \approx
0.25$ and $\nu \approx 0.3$, strength of interactions grows to a value that is
higher than the one given by default but soon starts to decrease as the
simulation runs. For FOSISS network, utility and strength of interactions fall
is quite steep but smooth, without sharp cuts. Between $\nu \approx 0.23$ and
$\nu \approx 0.28$, utility and strength of interactions start at their lowest,
but there is a slight increase in both of them.

@@

\includegraphics[scale=0.28]{images/erdos_trust_nus_1.png} & \includegraphics[scale=0.28]{watts_trust_nus_1.png} \\
\includegraphics[scale=0.28]{images/barabasi_trust_nus_1.png} & \includegraphics[scale=0.28]{fosiss_trust_nus_1.png}
\end{tabular}
\caption{Strength of interactions $w$ dynamics as a function of carrying capacity $\nu$ and time.}\label{trust}
\end{figure}

@@

Central to our argument is the differences in utility and strength of
interactions distribution at the end of the simulation, for every topology. We
found that utility distribution for the \textit{FOSISS} network, resembles quite
accurately to the distribution of utility in the Barab\'asi-Albert
network.\\

@@

Regarding strength of interactions distribution, the Erd\"{o}s-R\'enyi random
network displays a normal distribution of strength of interactions, as
expected. Again strength of interactions values are highly influenced by the
corresponding degree distribution (Figure 7A). However, in the Watts-Strogatz
network topology (Figure 7B), the strength of interactions distribution is a
highly asymmetric bimodal, with a really-low frequency mode of low strength of
interactions and a highly probability mode for high strength of interactions. A
possible explanation for this phenomenon is that under network topologies
maximizing inter-node communication (by minimizing the average distance between
nodes) such as the Watts-Strogatz, strength of interactions is favored both
among the cooperators (constituting the majority of players) and the
defectors. \\

The Barab\'asi-Albert network (7C) presents also a symmetric unimodal
distribution with values higher (on average) than those of the Erd\"{o}s-R\'enyi
random network, this may be the effect of increased communication due to more
efficient network navigability. Interestingly, the network corresponding to the
real FOSISS collaborations (7D) is an asymmetric unimodal distribution in which
moderate to high values of strength of interactions are more likely. We 
hypothesize that this effect is also due to the communication properties of the
network.

@@

\includegraphics[scale=0.28]{images/erdos_trust.png} & 
\includegraphics[scale=0.28]{images/watts_trust.png} \\
\includegraphics[scale=0.28]{images/barabasi_trust.png} & 
\includegraphics[scale=0.28]{images/fosiss_trust.png}
\end{tabular}
\caption{Strength of interactions distribution on different topologies. A. Random, Erd\"{o}s-R\'enyi network displays a normal-like
  distribution. B. Watts-Strogatz network strength of interactions distribution is bimodal and highly skewed to the right. C.
  Barab\'asi-Albert network strength of interactions distribution is also a normal-like curve. D. FOSISS, biomedical researchers
  collaboration network distribution of strength of interactions is skewed to
  the right.}\label{histo_trust}  
\end{figure}

@@

{\color{red}In this work we have analyzed the influence of parameters given by the underlying social structure of a science
collaborative network on collective strength of interactions and utility dynamical behavior, based on a class of iterated
PD and coordination games. Such parameters include mainly local and global connectivity like the degree centrality and
average clustering coefficient, as well as communication patterns.}\\

{\color{red}Under the assumptions given by the model, we were able to notice
  that, in general, communication within the social
collaborative networks has a positive correlation with average strength of interactions between the individuals partaking in the games and
also with the global collective utility (given by the sum of the individual payoffs). The better the
communication among players, the higher the strength of interactions and the utility leading to an optimized functioning of the whole
\emph{scientific collaboration system}. This is an important result that may be useful for scientific policy planning
and may set a foundation for the optimal use of social networks in scientific collaboration as a means to improve the
relationships among collaborating peers and thus ultimately the performance of research systems. 
We obviously need more qualitative work in order to validate these results from a sociological and anthropological
perspective.\\ %Furthermore, such validation is necessary because, as we said at
             %the beginning of this work, \emph{strength of interactions} can
             %have many different meanings and implementations in the real
             %world. The same we can say about \emph{utility} and most of all,
             %about the very idea of the game we explored here since ``playing
             %the game'' must have so, many different meanings among resarchers
             %and their communities.}\\

@@

{\color{red}The two main results of our model are the smoothness in the phase transition-like behavior for different parameters and the distribution of utility and strength of interactions in the FOSISS network compared to other networks and their topologies. We are certain that the particular structure of FOSISS network is playing a central role in the results and because of that we would like to discuss it a little bit further.}\\

@@

{\color{red}As is common among scientific communities in biomedical research, PIs play a central role in the network. Strong PIs and
well connected groups seem to be somehow responsible for the high levels of strength of interactions and the centralization of utility in
our simulation. As for what seems to be phase transitions, in the case of FOSISS networks, these are smoother than
those in the other networks with different topologies, even for those with a small-world topology. We think that this
behavior is also the result of the hierarchical structure already mentioned. If this is true, strength of interactions is first lost in the
edges that link diffrent goups and then in the edges that connect members of the groups. Connections between groups would not
be as dense as those inside the groups, which means that there would not be enough information of the behavior of one group regarding
its neighbors to constrain them as it seems to happen with individual researchers inside their communities. Nevertheless,
if values of the carrying capacity $\nu$ keep increasing and it becomes more
difficult to strengthen interactions, then strength of interactions begins to diminish inside groups.}\\

>>>>

\title{A game-theory modeling approach to utility
  and attachment dynamics in biomedical research social networks}
%\title{Two games modeling approach to collaboration dynamics in biomedical
%research social networks}

@@

  What would happen if researchers in a community were given the same
  amount of resources and then set free to interact among each other?
  How would these resources be distributed in the population after
  many interactions? The final result may be determined by the
  researchers' social and political power, an by their
  prestige. Nevertheless, all these attributes depend, to a certain
  degree, on the research collaboration social network in which
  scientists are embedded. Here we develop a model based on the
  Prisoner's Dilemma implemented on a collaboration network. With this
  model we explore the distribution of resources and the dynamics of
  the strengthening or weakening of collaboration connections among
  partners that we call attachment, as scientists cooperate or defect in order
  to gain access to others' resources. The network is based on data about projects
  that were awarded with a grant. In this sense, we assume that
  resources are funds, equipment and time. We define Utility as how
  much access a researcher has to the resources of his collegues.  We
  tested our simulation on the real biomedical research network and
  compared the results with an Erd\"{o}s-Reny\'i, a Watts-Strogatz
  small-world and Barab\'asi-Albert topologies. Different topologies
  display different utility and attachment
  distributions. Moreover, the distribution of utility and attachment in the
  researchers network is similar to that of 
  Barab\'asi-Albert and Watts-Strogatz topologies, respectively. We
  believe that utility distribution in the researchers network
  suggests that there are socio-cultural mechanisms governing the
  network that produce an asymmetric distribution of resources. The
  high distribution of strong connections might reflect some sort of
  subordination among researchers by which they are morally obliged to
  cooperate by the same socio-cultural mechanisms. The range around
  the threshold that regulates the decision to cooperate or defect
  according to the agent's historical balance between utility and
  strength of collaborative relationships and carrying capacity of the
  system is small, suggesting that there is a region in which a phase
  transition takes place from a population of cooperators to a
  population of defectors. Simulations like this may help to develop
  science policies to promote fair distribution of resources.

@@

%\textbf{Notes:}

%\begin{itemize}

@@

%\item We change utility for fitness. Utility seems a better concept in order to
  %reflect players' preferences.

@@

%\item Choosing a strategy depends on the balance $\eta'$ equation --look at page
 %%  9. This
%   equation states that the probability of an agent to cooperate increases as: a)
%   its utility increases --as it moves away from 0, or; b) The strength of the
%   connection with its peers increases --as it moves away from 0, or; c) both
%   increases.

%   $\eta_i = \frac{\langle f_i \rangle + \langle w_{ij} \rangle _j}{2}$\\

% \item{En la pagina 9 decimos que ``In order for the agent to choose to cooperate
%   or defect, a degree of willingness to cooperate is assigned to each agent.'' No estoy
%   seguro que ``degree of willingness to cooperate'' sea la mejor manera de
%   decirlo, mi duda está en la palabra ``degree'', por lo que cambi\'e la
%   oraci\'on a: The probability for an agent to cooperate or defect depends on a
% number that referes to a historical balance between average utility and the average strength of
% the connections with its neigbors. }

% \end{itemize}

@@

After each game, the agent adds-up utility ($u$), which is the sum of the
pay-offs following the PD matrix. A pair of neighbors will add-up to the strength of
their connection ($w_{ji}$) as they coordinate or anti-coordinate, being $w$
also cumulative. We measure global utility and connection strength for the whole
network. Global utility $U$ is the sum of all individual utilities and global
attachment or $W$ is the sum of every pair of agents' links
$w$. Metaphorically, the attachment can be thought as some sort of ``trust''.\\

{\color{red} It should be noted that the same actions or behaviors work for
  both $u$ and $w$. There are two reasons for this decision in the design of the
model. The most general one is that we believe that in the real world,
actions such as cooperating and defecting affect the attachment
among people. The second one is that we think that selfishly maximizing resources and
strenghting relationships are \emph{opposing forces} acting on the same set of
behaviors. The actions of an agent imply a trade-off in which defecting may
increase its utility at the expense of its collaborative relationships. If
collaborators have nourished their relationships, they might be strong enough to
endure occasional defection. Cooperating may build up relationships but it can be
expensive for the player.}

@@

The probability for an agent to cooperate or defect depends on a
number ($\eta$) that referes to a historical balance between average utility and
the average attachment with its neigbors.   {\color{red} This is so
  because we assume that whatever the result in utility or attachment, as long
  as one of them increases, the player will be confident in the strategy
  followed so far}.\\

@@

We ran two different experiments. In the first we simulated different values of
carrying capacity $\nu$. With this experiment we were able to see how the number of
cooperators, utility, attachment among agents and the ratio of
shifting state population would change in the range of the carrying
capacity. For this experiment the states of the agents were the same at
initialization, for all values of the carrying capacity. Since the model is
deterministic, it will return the same result if run under the same
conditions.\\

@@

In this section we present the main results of the study, namely the topological
structure of the underlying network models, the dynamics of the games under
different parameters and network topologies and the distribution of utility and
attachment resulting of playing the games in all the different scenarios
considered, including the real FOSISS network.\\

@@

Other results to report are those of the dynamics of different
variables as the carrying capacity $\nu$ changes. The  most salient
result is that for $\nu$ between $0.19$ and $0.24$, there is an
apparent phase transition in all different topologies and for all the
different variables. Nevertheless, it is worth noticing that the shape
of the phase transition is different according to the topology of the
network at stake. When $\nu$ is between $0.0$ and $0.2$, that is, when there is
no space or a very short one for suspiciousness, all agents cooperate, when
\textit{carrying capacity} is above $0.25$, all agents defect. \textit{Utility},
\textit{attachment}, and \textit{changing state population}
replicate that same behavior for the same limits. \\

@@

Utility and attachment dynamics under different $\nu$ emulate the
phase transition found before. In \textbf{Figures 3 and 4}, it can be observed
that there is a drop in utility and attachment according to the
drop in the number of cooperators. Erd\"{o}s-Reny\'i and Barab\'asi-Albert
networks are quite similar in the way these variables fall in two steps, the
first one at $\nu \approx 0.2$ and the next one at $\nu \approx 0.23$. The fall
is sharper still in the Barab\'asi-Albert topology. Utility and attachment phase
transition in Watts-Strogatz network is significantly more 
staggered compared to the former networks. In the case of utility, there is a
region in the limits of $\nu \approx 0.25$ and $\nu \approx 0.3$, before utility
goes to $0$, in which it remains low but stable over time. In general,
attachment follows the same pattern as utility but in the same $\nu \approx 
0.25$ and $\nu \approx 0.3$, attachment grows to a value that is
higher than the one given by default but soon starts to decrease as the
simulation runs. For FOSISS network, utility and attachment fall
is quite steep but smooth, without sharp cuts. Between $\nu \approx 0.23$ and
$\nu \approx 0.28$, utility and attachment start at their lowest,
but there is a slight increase in both of them.

@@

\includegraphics[scale=0.28]{images/erdos_trust_nus_1.png} & \includegraphics[scale=0.28]{watts_trust_nus_1.png} \\
\includegraphics[scale=0.28]{images/barabasi_trust_nus_1.png} & \includegraphics[scale=0.28]{fosiss_trust_nus_1.png}
\end{tabular}
\caption{Attachment $w$ dynamics as a function of carrying capacity $\nu$ and time.}\label{trust}
\end{figure}

@@

Central to our argument is the differences in utility and attachment
distribution at the end of the simulation, for every topology. We 
found that utility distribution for the \textit{FOSISS} network, resembles quite
accurately to the distribution of utility in the Barab\'asi-Albert
network.\\

@@

Regarding strength of interactions distribution, the Erd\"{o}s-R\'enyi random
network displays a normal distribution of attachment, as
expected. Again attachment values are highly influenced by the
corresponding degree distribution (Figure 7A). However, in the Watts-Strogatz
network topology (Figure 7B), the attachment distribution is a
highly asymmetric bimodal, with a really-low frequency mode of low strength of
interactions and a highly probability mode for high attachment. A
possible explanation for this phenomenon is that under network topologies
maximizing inter-node communication (by minimizing the average distance between
nodes) such as the Watts-Strogatz, attachment is favored both
among the cooperators (constituting the majority of players) and the
defectors. \\

The Barab\'asi-Albert network (7C) presents also a symmetric unimodal
distribution with values higher (on average) than those of the Erd\"{o}s-R\'enyi
random network, this may be the effect of increased communication due to more
efficient network navigability. Interestingly, the network corresponding to the
real FOSISS collaborations (7D) is an asymmetric unimodal distribution in which
moderate to high values of attachment are more likely. We 
hypothesize that this effect is also due to the communication properties of the
network.

@@

\includegraphics[scale=0.28]{images/erdos_trust.png} & 
\includegraphics[scale=0.28]{images/watts_trust.png} \\
\includegraphics[scale=0.28]{images/barabasi_trust.png} & 
\includegraphics[scale=0.28]{images/fosiss_trust.png}
\end{tabular}
\caption{attachment distribution on different topologies. A. Random, Erd\"{o}s-R\'enyi network displays a normal-like
  distribution. B. Watts-Strogatz network attachment distribution is bimodal and highly skewed to the right. C.
  Barab\'asi-Albert network attachment distribution is also a normal-like curve. D. FOSISS, biomedical researchers
  collaboration network distribution of attachment is skewed to
  the right.}\label{histo_trust}  
\end{figure}

@@

{\color{red}In this work we have analyzed the influence of parameters given by the underlying social structure of a science
collaborative network on collective attachment and utility dynamical behavior, based on a class of iterated
PD and coordination games. Such parameters include mainly local and global connectivity like the degree centrality and
average clustering coefficient, as well as communication patterns.}\\

{\color{red}Under the assumptions given by the model, we were able to notice
  that, in general, communication within the social
collaborative networks has a positive correlation with average attachment between the individuals partaking in the games and
also with the global collective utility (given by the sum of the individual payoffs). The better the
communication among players, the higher the attachment and the utility leading to an optimized functioning of the whole
\emph{scientific collaboration system}. This is an important result that may be useful for scientific policy planning
and may set a foundation for the optimal use of social networks in scientific collaboration as a means to improve the
relationships among collaborating peers and thus ultimately the performance of research systems. 
We obviously need more qualitative work in order to validate these results from a sociological and anthropological
perspective.\\ %Furthermore, such validation is necessary because, as we said at
             %the beginning of this work, \emph{attachment} can
             %have many different meanings and implementations in the real
             %world. The same we can say about \emph{utility} and most of all,
             %about the very idea of the game we explored here since ``playing
             %the game'' must have so, many different meanings among resarchers
             %and their communities.}\\

@@

{\color{red}The two main results of our model are the smoothness in the phase transition-like behavior for different parameters and the distribution of utility and attachment in the FOSISS network compared to other networks and their topologies. We are certain that the particular structure of FOSISS network is playing a central role in the results and because of that we would like to discuss it a little bit further.}\\

@@

{\color{red}As is common among scientific communities in biomedical research, PIs play a central role in the network. Strong PIs and
well connected groups seem to be somehow responsible for the high levels of attachment and the centralization of utility in
our simulation. As for what seems to be phase transitions, in the case of FOSISS networks, these are smoother than
those in the other networks with different topologies, even for those with a small-world topology. We think that this
behavior is also the result of the hierarchical structure already mentioned. If this is true, attachment is first lost in the
edges that link diffrent goups and then in the edges that connect members of the groups. Connections between groups would not
be as dense as those inside the groups, which means that there would not be enough information of the behavior of one group regarding
its neighbors to constrain them as it seems to happen with individual researchers inside their communities. Nevertheless,
if values of the carrying capacity $\nu$ keep increasing and it becomes more
difficult to strengthen interactions, then attachment begins to diminish inside groups.}\\
